Multi-level modelling with Openflexo/FML - A contribution to the \mpc

### 2021-05-27 3-Introduction.tex

In order to tackle this problem, the \emph{multi-level} modeling paradigm have been introduced. Contrary to the strict approach, multi-level modeling advocates the use of a flexible number of levels as well as a flexibilization of the relations between them. This paradigm is gaining traction lately within the modeling community as evidenced by the contribution of many new multi-level modeling approaches and tools and the celebration of topic specific workshops\footnote{https://homepages.ecs.vuw.ac.nz/Groups/MultiLevelModeling/WebHome} and seminars~\citep{almeida2018multi}. Taking advantage of this vibrant state of affairs, and in order to foster discussion and enable comparison between competing approaches, a multi-level modeling challenge has been created. This paper is a response to the latest multi-level modeling challenge
which consists in providing a solution to the problem of specifying and enacting processes. Solutions to the \emph{Multi-level Process Challenge} must fulfill a number of requirements for a process representation defined at an abstract process-definition level and at various more concrete domain-specific levels, resulting in a multi-level hierarchy of related models. We provide a solution based on model federation~\citep{Golra2016-federation}.

@@

The rest of the paper is organized as follows. Section~\ref{sec:technology} presents our approach and the technology it relies on. Section~\ref{sec:analysis} is the analysis of the problem. We detail our model in section~\ref{sec:model} and show it satisfies the challenge requirements in section~\ref{sec:requirements}. We assess the modeling solution in section~\ref{sec:discussion}. Section~\ref{sec:relatedwork} presents the related work before the conclusion~\ref{sec:conclusion}.

>>>>

In order to tackle this problem, the \emph{multi-level} modeling paradigm have been introduced. Contrary to the strict approach, multi-level modeling advocates the use of a flexible number of levels as well as a flexibilization of the relations between them. This paradigm is gaining traction lately within the modeling community as evidenced by the contribution of many new multi-level modeling approaches and tools such as Melanee~\citep{melanee}, MetaDepth~\citep{metadepth}, MultEcore~\citep{multecore2016}, DeepTelos\citep{deeptelos2016} or DMLA~\citep{dmla2017}. Taking advantage of this vibrant state of affairs, and in order to foster discussion and enable comparison between competing approaches, a multi-level modeling challenge has been created. This paper is a response to the latest multi-level modeling challenge
which consists in providing a solution to the problem of specifying and enacting processes. Solutions to the \emph{Multi-level Process Challenge} must fulfill a number of requirements for a process representation defined at an abstract process-definition level and at various more concrete domain-specific levels, resulting in a multi-level hierarchy of related models. We provide a solution based on model federation~\citep{Golra2016-federation}.

@@

The rest of the paper is organized as follows. Section~\ref{sec:technology} presents our approach and the technology it relies on. Section~\ref{sec:analysis} is the analysis of the problem. We detail our model in section~\ref{sec:model} and show it satisfies the challenge requirements in section~\ref{sec:requirements}. We assess the modeling solution in section~\ref{sec:discussion}. Section~\ref{sec:relatedwork} presents the related work before the conclusion.%~\ref{sec:conclusions}.

### 2021-05-27 4-Technology.tex

FML is a language designed to define virtual models. A virtual model
is composed of a set of \emph{concepts}, while itself being a concept.
Hence, virtual models are structuring units while concepts are the
core entities. A concept has a set of \emph{roles} and
\emph{behaviors}. A parallel to object-oriented approach can be useful
to understand FML\footnote{Some aspects of FML do not exist in object oriented approach.}. A concept corresponds to a class, its roles to the
attributes of the class and its behaviors to the methods of the class.
These roles have types defining the kind of value the role will point
at runtime.
Whenever a type external to the federation space (from a TA or an
external model) is used, one needs to use a \emph{model slot}. A model
slot is a mediation entity, associated with a TA, in charge of giving
access to external elements of the corresponding technological space.

A \emph{technology adapter} (TA) is a reusable library that defines
connections between the FML execution engine and a particular
technological space. The model federation framework provides ways to
define TA.

We should stress that building a virtual model is purely conceptual,
model slots / TA are in charge of the technicalities. Additionally,
when a virtual model or a concept point to an external model, it comes
with an explicit interpretation of the external data. The virtual
model can hide information, transform elements and does not need to
cover the whole external metamodel.
% In our example, it ignores all other elements of the \smk metamodel.
We did not required TA in our proposal. The whole solution uses the
internal (FML) language.

FML is designed to define not only the structure of virtual models but
also to define the collection of actions an engineer can perform on
them. These actions are called \emph{business behaviors} and may be
compared to methods of object-oriented approach.
% The rename operation already cited for the \textsf{Goal-xls}
% federation is an example of such a behavior. One could also define
% operations to add or remove requirements, to refine relationships,
% \emph{etc.} The renaming operation is defined by a \texttt{rename}
% behavior in the concept \texttt{Requirement}. It uses the
% \texttt{setCell} action of the Excel TA and the \texttt{rnGoal} of
% the \smk TA.
When the FML execution engine runs a federation, it creates virtual
model instances containing concept instances. Some concept instances
are connected to external elements through model slot instances.
% The lower part of the figure~\ref{fig:space-ex} illustrates this
% runtime phase.

FML also supports another kind of behavior which can be described as
interactors. Such behavior is reactive and triggers when any of the
federated models evolves. Two scenarios are then possible: (1) an
action in the federation is propagated to the federated models or (2)
a modification is detected in a federated model and triggers a
behavioral interactor of the federation.
% Hence the double arrows in the links of Figure~\ref{fig:space-ex}.
This feature is useful when models have to be synchronised. We did not
exploit this possibility in the challenge.

>>>>

\begin{figure}[t]
    \centering
    \includegraphics[width=\columnwidth]{Figures/FMLCoreModel.pdf}
    \caption{The FML core metamodel}
    \label{fig:mm}
\end{figure}

FML simplified metamodel is provided in Figure~\ref{fig:mm}, it is designed to define virtual models. A virtual model
is composed of a set of \emph{concepts}, while itself being a concept.
Hence, virtual models are structuring units while concepts are the
core entities. A concept has a set of \emph{roles} and
\emph{behaviors}. A parallel to object-oriented approach can be useful
to understand FML\footnote{Some aspects of FML do not exist in object oriented approach.}. A concept corresponds to a class, its roles to the
attributes of the class and its behaviors to the methods of the class.
These roles have types defining the kind of value the role will point
at runtime.
Whenever a type external to the federation space is used, one needs to use a \emph{model slot}. A model
slot is a mediation entity in charge of giving
access to external elements using
a \emph{technology adapter}\footnote{It is a reusable library that defines
connections between the FML execution engine and a particular
technological space.}.

FML is designed to define not only the structure of virtual models but
also to define the collection of actions an engineer can perform on
them. These actions are called \emph{behaviors} and can either be called (like usual methods) or triggered by events. The reactive behaviors are mainly useful when models have to be synchronised. We did not
exploit this possibility in the challenge.

When the FML execution engine runs a federation, it creates virtual
model instances containing concept instances. Some concept instances
are connected to external elements through model slot instances.

### 2021-05-27 8-Discussion.tex

  Due to the fact our approach is level-blind, our solution could have had more
  or fewer levels depending on the variations of the use case. %As levels are
  %not specifically encoded,

@@

\noteJC{on a du tooling, un peu moins de formalisme}

Currently, there are no formalism to establish the semantics of the MLM
technique in our solution. \textcolor{red}{To develop…}

@@

By default, our tooling does not include a lot of automated verifications.
However, a user can add verification mechanisms by writing new behaviors
attached to concepts. Presently, the verifications rely on the users when they
build the metamodels and the models.

>>>>

  Due to the fact our approach is level-blind, our solution could have had more
  or fewer levels depending on the variations of the use case. However, the number of levels is very related to the problem, making the approach fixed. For example, given a requirement, an instance cannot become a concept easily.

@@

\noteJC{on a du tooling, un peu moins de formalisme}
-
Currently, there are no formalism to establish the semantics of the MLM
technique in our solution. \textcolor{red}{To develop…}

@@

Our tooling includes mechanisms to verify the models. First, syntactical consistency is realised by cardinality checks and by typing. Second, the constraints the users have written are continuously checked in order to

However, a user can add verification mechanisms by writing new behaviors
attached to concepts. Presently, the verifications rely on the users when they
build the metamodels and the models.
